---
title: "Homework 1"
---

```{r global_options, include=FALSE}
library(knitr)
library(ggplot2)
opts_chunk$set(fig.align="center", fig.height=4, fig.width=5.5)
```

**This homework is due on the deadline posted on edX. Please submit a .pdf file of your output and upload a .zip file containing your .Rmd file. Do NOT include your name or EID in your filenames.**

In this homework you will be working with the `chickwts` dataset built into R. This data set contains weight measurements of chicks fed on different food sources to assess their effectiveness on growth rate.

```{r}
head(chickwts)
```

**Problem 1:** Use ggplot to make a histogram of the `weight` column. Manually choose appropriate values for `binwidth` and `center`. Explain your choice of values in 2-3 sentences.

```{r}
summary(chickwts)
```

```{r}
ggplot(chickwts, aes(weight)) + 
  geom_histogram(
    binwidth = 20,
    center = 10,
    fill = 'blue',
    color = 'black')
```

*Your explanation goes here.*

Before selecting the *bindwidth* and *center*, I decided to look at the summary statistics of the dataset, which would help in evaluating my choices. Upon noticing that the weights range from ~100 to ~420, I experimented with different multiples of 10, and found that using **20** as my binwidth provided interesting insights - there are several peaks and troughs amongst the distribution of weights. Finally, I selected my center as **10** because it presents bins in a clean fashion (i.e. from 100 to 120, 200 to 220, etc.).


**Problem 2:** Modify the plot from Problem 1 to show one panel per feed. Hint: Use `facet_wrap()`.

```{r}
ggplot(chickwts, aes(weight)) + 
  geom_histogram(
    binwidth = 20,
    center = 10,
    fill = 'blue',
    color = 'black') +
  facet_wrap(vars(feed))
```

